---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BulsinkBxG

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/pbulsink/BulsinkBxG/workflows/R-CMD-check/badge.svg)](https://github.com/pbulsink/BulsinkBxG/actions)
<!-- badges: end -->

The goal of BulsinkBxG is to share a basic Expected Goals (xG) model for NHL data from the year 2011 and onward. Expected Goals are generated from shot data, taking into consideration details about the shot like distance and angle to the net, whether the shot is a rebound, etc. By modelling across seasons of data, we're able to assign a value of how likely each shot was to becoming a goal. This in turn can show teams that have performed well (i.e. produced lots of high-quality shot attempts) despite scoring few goals. Expected goals can also be used to highlight players with good (or bad) impacts on their team (either producing xG above or below average, or having more or less xG against than their teammates).

## Installation

You can install the released version of BulsinkBxG from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pbulsink/BulsinkBxG")
```

## Example

The default xG models for each playing year from 2011 onward are included in this package. They are produced using up to the past 3 seasons' data and validated against the target season. Model are loaded using the `load_season_model()` function.

```{r example}
library(BulsinkBxG)

#load the model for analyzing the 2012 season data:
model_2012 <- load_season_model(2012)

#load the pbp data for 2012:
pbp_2012 <- load_season_pbp(2012)

#filter pbp data to (fenwick) shot attempts only:
pbp_fenwick_2012 <- pbp_2012 %>%
  dplyr::filter(.data$event %in% c('Shot', 'Missed Shot', 'Miss', 'Goal'))

#predict an xG value for each shot attempt:
pbp_fenwick_2012$xG <- predict(model_2012, new_data = pbp_fenwick_2012, type = "prob")$.pred_goal
```

We can plot the xG expected by location to see the effect of distance and angle:

```{r plot_xg_xy}
library(ggplot)
avg_xG_by_coord <- pbp_fenwick_2012 %>% group_by(.data$x_coord, .data$y_coord) %>%
    summarise(xg_mean = mean(xG))

ggplot(avg_xG_by_coord, aes(x_coord, y_coord, fill = xg_mean)) + geom_raster() +
    scale_fill_gradient(low = 'blue', high = 'red')+
    geom_vline(xintercept = 0, color = 'red') +
    geom_vline(xintercept = 25, color = 'blue') +
    geom_vline(xintercept = 88, color = 'red') +
    xlab('X Coordinates') + ylab('Y Coordinates') +
    labs(title = 'Average xG Value by Coordinate')
```


# Building your own xG model

This package contains the required tools to download, parse, and reprocess the NHL's Play by Play (pbp) and shift data to build your own data set. This can be used to build your own xG model or other analytics.

## Data Storage

The package also contains the code to download and process the NHL's Play by Play (pbp) & shift data to train your own version of the model. The data is stored (by default) in the folder `~/NHLpbp` on your local device. If you wish to change this location, set the value of the option `BulsinkBxG.data.path` either interactively or in your .RProfile by the following command:

```option("BulsinkBxG.data.path" = [new path])```

## Downloading the Data

The first step is to run the following code to download and process the pbp and shift data, provinding a season:
```{r pbp, eval=FALSE}
pbp<-process_season_pbp(2011)
```

Repeat this for each season from 2011, or in short use `purrr`: 

```
purrr::walk(2011:2021, function(x) process_season_pbp(x))
```

In future sessions, the processed data will be loaded instead of downloaded and parsed (saving significant time). 

## Preparing Model Features

Much of the data used by the expected goals model is not present directly in the pbp file. Process each season's data using `load_prepped_data()` or, again with purrr:

```
purrr::walk(2011:2021, function(x) load_prepped_data(x))
```

In future sessions, the processed data will be loaded instead of reprocessed (saving significant time).  

## Building a model

Building the xG model itself can be performed by running `build_model()`. Note that a 2011 model would be built using the 2011 season data, which may not be useful. 

```
purrr::walk(2011:2021, function(x) build_model(x))
```

Note that this will take a very long time. If you plan to use the default model, please just load the model using `load_season_model()` function.

## GPU Usage for XGBoost
If you have XGBoost installed with GPU configuration, and want to train using the GPU, set the `BulsinkBxG.xgboost.gpu` option to `TRUE`. This option is `FALSE` by default. More information on XGBoost GPU can be found at [https://xgboost.readthedocs.io/en/stable/gpu/index.html](https://xgboost.readthedocs.io/en/stable/gpu/index.html), with installation guide in [the docs](https://xgboost.readthedocs.io/en/stable/install.html#r). 

# Model Performance Metrics

This model uses a custom metric to compare performance of the predicted xG to the actual goals. In addition to using accuracy, log loss and ROC AUC, a custom metric called Total Error Sum (TES) has been developed. The TES metric compares the total sum of goals in the predictor, and the total sum of the xG values. If the TES is positive, there are more goals than expected goals in the season. If negative, there are more xG than goals.

The model provides the following performance by each year in the logistic (gbm) version:

| Season | ROC AUC | Log Loss | (Sum (G) - Sum(xG))^2 |
|--------|---------|----------|-----------------------|
| 2011   | 0.907   | 0.165    |  51.28                |
| 2012   | 0.914   | 0.152    |  91.58                |
| 2013   | 0.919   | 0.149    | 116.09                |
| 2014   | 0.918   | 0.148    |  37.43                |
| 2015   | 0.917   | 0.148    | 116.68                |
| 2016   | 0.914   | 0.149    | 132.59                |
| 2017   | 0.914   | 0.151    | 226.04                |
| 2018   | 0.908   | 0.159    | 230.93                |
| 2019   | 0.908   | 0.162    | 321.33                |
| 2020   | 0.911   | 0.158    | 206.54                |
| 2021   | 0.906   | 0.162    |  46.92                |

Similarly, the XGBoost version's performance metrics are shown here: 

| Season | ROC AUC | Log Loss | Sum (G) - Sum(xG)^2 |
|--------|---------|----------|---------------------|
| 2011   |   --    | --       |  --                 |
| 2012   | 0.89737 |          | 151.2               |
| 2013   |         |          |                     |
| 2014   |         |          |                     |
| 2015   |         |          |                     |
| 2016   |         |          |                     |
| 2017   |         |          |                     |
| 2018   |         |          |                     |
| 2019   |         |          |                     |
| 2020   |         |          |                     |
| 2021   | --      | --       | --                  |

# Code of Conduct

Please note that the BulsinkBxG project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
